---
title: "Hypothesis testing for proteomics with h0testr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hypothesis testing for proteomics with h0testr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment="#>")
```

## Why this package

`h0testr` analyzes non-negative continuous expression matrices, such as those produced by
mass-spec-based proteomics.

In proteomics there are many methods available for inter-sample normalization, feature
aggregation, missing value imputation, and hypothesis testing. Many of the same methods are
also applicable to, or have been borrowed from, other fields of expression data analysis,
such as RNA-seq. Many of these methods have parameters whose settings can affect the quality
of the results.

The package was initially developed as a test platform for evaluating the performance of
different workflow configurations and parameter settings. To that end it provides a uniform
higher-level interface to a wide variety of methods, and a wrapper-based building-block
design that abstracts away differences between their native interfaces. That design supports
two ways of working: run a whole configured workflow with one call, or call the individual
steps yourself.

## Installation

The default workflow needs only `limma`, `edgeR` and `MsCoreUtils` beyond base R, so that is
all the install pulls in. Every alternative normalization, imputation and testing engine is
optional: ask for one whose package is missing and the run stops with the install command for
that package.

```{r install, eval=FALSE}
## some dependencies are on Bioconductor, so add its repositories first:
install.packages(c("BiocManager", "remotes"))
options(repos=BiocManager::repositories())

install.packages("h0testr")
help(package="h0testr")
```

Optional engines, installed only if you plan to use them. With the repositories set as
above, `install.packages()` finds the Bioconductor ones too:

```{r install_optional, eval=FALSE}
install.packages(c(
  "glmnet",                        ## impute_method "glmnet"
  "imputeLCMD",                    ## impute_method "min_det", "min_prob", "qrilc"
  "missForest",                    ## impute_method "missforest"
  "randomForest",                  ## impute_method "rf"
  "DEqMS",                         ## test_method "deqms"
  "impute",                        ## impute_method "knn"
  "pcaMethods",                    ## impute_method "bpca", "ppca", "svdImpute", "lls"
  "vsn",                           ## normalization_method "vsn"
  "proDA", "SummarizedExperiment", ## test_method "proda"
  "msqrob2", "QFeatures",          ## test_method "msqrob", "msqrob_agg"
  "lme4", "lmerTest"               ## test_method "msqrob_agg", "prolfqua_lmer"
))

## test_method "prolfqua" and "prolfqua_lmer"; github only:
remotes::install_github("fgcz/prolfqua")
```

## A first run

Everything about a run lives in one list, which `new_config()` returns pre-filled with
defaults. You edit the keys your experiment needs and hand the list to `run()`.

The example data shipped with the package is 657 peptides belonging to 100 genes, measured
over 12 observations in two groups (`grp`, `ctl` versus `trt`) balanced across `sex`. It is
small enough to run in a second and has the same shape as real input.

```{r config}
config <- h0testr::new_config()

## where the three input files are, and where output goes:
config$dir_in <- system.file("extdata", package="h0testr")
config$dir_out <- tempdir()
config$feature_file_in <- "features2.tsv"
config$sample_file_in <- "samples2.tsv"
config$data_file_in <- "expression2.tsv"

## how the three files cross-reference each other:
config$feat_id_col <- "pep"      ## column of features file matching rownames of expression
config$gene_id_col <- "gene"     ## column of features file to aggregate peptides into
config$obs_id_col <- "obs"       ## column of samples file matching colnames of expression
config$sample_id_col <- "obs"    ## biosample id; same as obs_id_col when no technical reps

## what to test:
config$frm <- ~grp + sex         ## model to fit
config$test_term <- "grp"        ## term of frm to test
config$reference_levels <- c(grp="ctl", sex="F")   ## every factor needs its reference level

## how to process:
config$normalization_method <- "RLE"
config$impute_method <- "unif_sample_lod"
config$impute_quantile <- 0
config$test_method <- "trend"
config$n_features_min <- 10       ## the default of 1000 is too big for this small demo

config$verbose <- FALSE           ## the log is long; this vignette silences it
config$log_file <- tempfile()
```

`run()` loads the data, walks the steps named in `config$run_order`, and tests:

```{r run}
result <- h0testr::run(config)
names(result)
```

```{r run_hits}
head(result$standard)
```

`result$standard` is the hit table in a standardized format, the same columns whatever
engine produced it, so that methods can be compared. `result$original` is the table as the
underlying package returned it, `result$fit` is the fitted model where the engine returns
one, and `result$state` is the processed data the test saw.

```{r run_extras}
dim(result$state$expression)              ## 100 genes after aggregation, 12 observations
sum(result$standard$adj_pval < 0.05)      ## hits at 5% FDR
```

## Input format

Inputs are three tab-delimited text files in `config$dir_in`, named by
`config$feature_file_in`, `config$sample_file_in` and `config$data_file_in`. Examples are in
`system.file("extdata", package="h0testr")`.

**`data_file_in`** holds non-negative **raw** (not log-transformed) expression, rows being
features (precursor, peptide, protein group or gene group) and columns being observations
(fraction, technical replicate or sample). Both `0` and `NA` are treated as missing. The file
is assumed to have been written from R with row and column names, so its first line has one
fewer field than the rest. **Row and column names must be unique.**

```{r show_expression}
exprs_file <- file.path(system.file("extdata", package="h0testr"), "expression2.tsv")
exprs <- as.matrix(read.table(exprs_file, header=TRUE, sep="\t", quote="", as.is=TRUE))
exprs[1:4, 1:3]
```

**`feature_file_in`** holds feature metadata, with column names and no row names. The column
named by `config$feat_id_col` must match `rownames(expression)`, so that after loading
`all(features[, feat_id_col] == rownames(expression))`. The column named by
`config$gene_id_col` holds the protein-group or gene-group ids that features are aggregated
into. Any remaining columns are carried through and appear as annotation in the results.

```{r show_features}
feats_file <- file.path(system.file("extdata", package="h0testr"), "features2.tsv")
feats <- read.table(feats_file, header=TRUE, sep="\t", quote="", as.is=TRUE)
head(feats, 3)
```

**`sample_file_in`** holds observation metadata, with column names and no row names. It needs
the column named by `config$obs_id_col`, matching `colnames(expression)`; the column named by
`config$sample_id_col`, which is not unique when there is technical replication; and every
variable referred to in `config$frm`.

```{r show_samples}
samps_file <- file.path(system.file("extdata", package="h0testr"), "samples2.tsv")
samps <- read.table(samps_file, header=TRUE, sep="\t", quote="", as.is=TRUE)
head(samps, 3)
```

## Configuration

`new_config()` returns the defaults, with the most frequently changed settings towards the
top. **At a minimum** edit `frm`, `test_term` and `reference_levels`: the defaults are almost
certainly wrong for your experiment.

```{r new_config}
str(h0testr::new_config(), list.len=8, give.attr=FALSE)
```

`report_config()` prints a run's settings, and `check_config()` validates them and is called
at the start of each step.

### The formula and its reference levels

`config$frm` accepts `+`, `:` and `*`, and `~0 +` to drop the intercept. `I()`, `poly()`,
splines and random-effect terms `(1|x)` are out of scope. `config$test_term` names one term
of `config$frm`; alternatively `config$contrast` names a weighted sum of coefficients to test
instead, such as `"grpb - grpc"`.

Only the reference level of each factor is declared, in `config$reference_levels`; the
remaining levels are sorted after it. So a variable not named there is continuous if it is
numeric, and an **error** if it is character, since its reference level, and hence the
meaning of its coefficients, would otherwise be decided by locale-dependent sorting. Logical
variables need no declaration, being always ordered `FALSE`, `TRUE`. The resolved ordering is
written to the log and returned in `config$factor_levels`.

### Output files

Setting `config$save_state <- TRUE` writes each intermediate state to `config$dir_out` as
three tab-delimited files -- expression, features, samples -- named by `config$data_mid_out`,
`config$feature_mid_out` and `config$sample_mid_out`, with `config$suffix_out`. The numeric
prefix reflects the order the steps ran in, which `config$run_order` determines. With the
defaults:

```
1.initial.*          ## covariates reduced to what is needed downstream, filter stats added
2.prepped.*          ## uninformative features and observations dropped; permuted if asked
3.normalized.*       ## inter-sample normalized and log2- or glog2-transformed
4.combined_replicates.*   ## technical replicates aggregated to samples
5.combined_features.*     ## features aggregated to protein or gene groups
6.filtered.*         ## filtered for min samples/feature and features/sample
7.imputed.*          ## missing values imputed
8.results.tsv        ## test results
```

## Workflow overview

A run always begins by loading data and ends by testing hypotheses. The steps in between are
named by `config$run_order`, whose default gives:

```{r run_order}
h0testr::new_config()$run_order
```

1. **Load data.** Read the three files, parse, reformat, optionally permute, and check the
   three for consistency with each other.
2. **Normalize** between samples, by `config$normalization_method`. This includes a
   `log2(x+1)` transform or equivalent, unless the method is `"none"`.
3. **Combine technical replicates**, by median.
4. **Combine features** into gene or protein groups, by `config$feature_aggregation`. Skip
   this step for `test_method` in `c("deqms", "msqrob", "msqrob_agg", "prolfqua_lmer")`,
   which take feature-level input and report one row per gene themselves.
5. **Filter** features and observations, by `config$n_samples_min`,
   `config$n_features_min`, `config$df_resid_min` and `config$estimability`.
6. **Impute** missing values by `config$impute_method`. Both `0` and `NA` are missing.
7. **Test** the null hypothesis that the effect of `config$test_term` is `0`, by
   `config$test_method`; or, when `config$contrast` is set, that the weighted sum of
   coefficients it names is `0`.

## Running one step at a time

Sometimes you begin with R objects rather than files, or want to run the steps yourself. The
state is a plain list of three elements named `expression`, `features` and `samples`, with
the expression a numeric matrix whose dimnames are kept in sync with the two data frames
throughout.

```{r state}
state <- list(expression=exprs, features=feats, samples=samps)
is.matrix(state$expression)
all(rownames(state$expression) == state$features$pep)
all(colnames(state$expression) == state$samples$obs)
```

The same `config` drives the steps individually. `init_state()` and the steps that follow
return a list with both `$state` and `$config`, since a step may learn something the next one
needs -- `normalize()`, for instance, records that the data is now log-transformed.

```{r stepwise}
out <- h0testr::init_state(state, config)
out$state <- h0testr::add_filter_stats(out$state, out$config)
out$state <- h0testr::prefilter(out$state, out$config)
out$state <- h0testr::permute(out$state, out$config)
out <- h0testr::normalize(out$state, out$config)
out <- h0testr::combine_replicates(out$state, out$config)
out <- h0testr::combine_features(out$state, out$config)
out <- h0testr::filter_state(out$state, out$config)
out <- h0testr::impute(out$state, out$config)
result2 <- h0testr::test_h0(out$state, out$config)

head(result2$standard, 3)
```

Individual engines can also be called directly, bypassing the dispatch: `normalize_edger()`,
`impute_unif_sample_lod()`, `test_trend()` and their siblings. See their help pages.

## Reading the results

`result$standard` has one row per tested feature, sorted by p-value:

| column | meaning |
|---|---|
| `feature` | the feature tested; a gene group for the gene-level methods |
| `expr` | average expression over the observations where the feature was seen |
| `logfc` | estimated effect size; see below |
| `stat` | the test statistic the engine reported |
| `lod` | log-odds of differential expression; only `trend` and `voom`, else `NA` |
| `pval` | raw p-value |
| `adj_pval` | p-value adjusted for multiple testing |

What `logfc` holds depends on how many design-matrix columns `config$test_term` resolves to,
which is the same for every row of one result:

- **One column:** that coefficient, signed. For a two-level factor this is the difference
  between its levels on the scale of the expression, so a log fold change when the input is
  log-transformed. For a **continuous** covariate it is the change **per unit** of that
  covariate, so its size depends on the units the covariate is recorded in -- an effect per
  month is a twelfth of the same effect per year. Thresholds on `abs(logfc)` have to be
  chosen with that in mind, though rankings and p-values are unaffected.
- **More than one column:** a joint test has no single contrast to report, so what is
  reported is the total swing -- the range, over the observations, of the fitted contribution
  of the terms under test. It is unsigned, and on the same scale as the single-column case.
- **`config$contrast`:** the weighted sum of the coefficients it names, signed. A contrast is
  one degree of freedom however many coefficients it weights, so unlike a joint test it
  always has one number to report.

`result$original` carries the engine's own columns, plus the feature annotation from
`feature_file_in` and the statistics the pipeline added:

```{r original}
colnames(result$original)
```

## Choosing methods

Each of the three lists below can be printed from R, and each names the `config` key that
selects from it.

```{r methods}
h0testr::normalize_methods()
h0testr::impute_methods()
h0testr::test_methods()
```

Reasonable starting points: for DIA, `normalization_method="RLE"` with
`impute_method="unif_sample_lod"` and `impute_quantile=0.05`; for DDA,
`normalization_method="quantile"` with `normalization_quantile=0.75`,
`impute_method="unif_sample_lod"` and `impute_quantile=0`.

### Normalization

Every method other than `"none"` leaves the data on a log-like scale. Results are
`log2(x+1)`-transformed afterwards, with three exceptions: `"none"` returns the input
untouched on whatever scale it arrived; `"vsn"` is arsinh-scaled by its own code and so is
not transformed again; and `"loess"` has its input transformed *before* normalization rather
than after, because cyclic loess is an additive correction that on raw intensities returns
negative fitted values for the smallest measurements, whose `log2(x+1)` would be `NaN`.

| method | what it does |
|---|---|
| `cpm` | counts per million: `multiplier * intensities / sum(intensities)` per sample |
| `div.mean` | subtract the observation's mean across features; `MsCoreUtils::normalize_matrix()` |
| `div.median` | subtract the observation's median across features; `MsCoreUtils::normalize_matrix()` |
| `loess` | cyclic loess; `limma::normalizeCyclicLoess()`, span `config$normalization_span` |
| `log2` | `log2(x+1)` and nothing else |
| `max` | divide each feature by its max; `MsCoreUtils::normalize_matrix()` |
| `none` | unchanged |
| `qquantile` | classical quantile normalization; `limma::normalizeQuantiles()` |
| `quantile` | divide each sample by its `config$normalization_quantile` quantile |
| `quantiles.robust` | robust quantile normalization; `MsCoreUtils::normalize_matrix()` |
| `RLE` | relative log expression; `edgeR::calcNormFactors()` then `edgeR::cpm()` |
| `sum` | divide each feature by its sum; `MsCoreUtils::normalize_matrix()` |
| `TMM` | trimmed mean of M-values; `edgeR::calcNormFactors()` then `edgeR::cpm()` |
| `TMMwsp` | TMM with singleton pairing; may beat TMM when many zeros are present |
| `upperquartile` | upper-quartile scaling; `edgeR::calcNormFactors()` then `edgeR::cpm()` |
| `vsn` | variance-stabilizing transformation; `limma::normalizeVSN()` |

`RLE` needs at least one feature measured in every observation, since that is where it takes
its reference from; that is not always so in data with high missingness, and `TMM`, `TMMwsp`
and `upperquartile` do not need one.

### Imputation

Both `0` and `NA` are treated as missing. `glmnet` and `rf` are slow and may take days on
large datasets.

| method | what it does |
|---|---|
| `bpca` | Bayesian PCA; `pcaMethods::pca(method="bpca")` |
| `glm_binom` | GLM of `p(missing) ~ log2(intensity)`, logit link; draw from the fitted density |
| `glmnet` | model a feature's missing intensity on the other features; `glmnet::cv.glmnet()` |
| `knn` | k-nearest neighbors; `impute::impute.knn()` |
| `lls` | linear least squares; `pcaMethods::llsImpute()` |
| `loess_logit` | as `glm_binom`, but a local linear fit of `logit(p(missing))` |
| `min_det` | a low quantile of the observed signal; `imputeLCMD::impute.MinDet()` |
| `min_prob` | draws from a normal near the limit of detection; `imputeLCMD::impute.MinProb()` |
| `missforest` | the missForest algorithm; `missForest::missForest()` |
| `none` | leave missing; features or samples may then be dropped by the test |
| `ppca` | probabilistic PCA; `pcaMethods::pca(method="ppca")` |
| `qrilc` | quantile regression; `imputeLCMD::impute.QRILC()` |
| `rf` | as `glmnet`, with `randomForest::randomForest()` |
| `rnorm_feature` | per feature, `rnorm(1, mean=m, sd=s * config$impute_scale)` |
| `sample_lod` | per sample, the minimum observed value in that sample |
| `svdImpute` | the svdImpute algorithm; `pcaMethods::pca()` |
| `unif_global_lod` | `runif(1, 0, q)`, `q` the `impute_quantile` of per-feature minima |
| `unif_sample_lod` | `runif(1, 0, q)`, `q` the `impute_quantile` of a sample's observed values |

`impute_method="none"` is not compatible with `test_method` in `c("msqrob", "voom")`, which
cannot take missing values.

### Testing

| method | what it does |
|---|---|
| `deqms` | `limma::lmFit()` -> `eBayes()` -> `DEqMS::spectraCounteBayes()` -> `topTable()` |
| `lm` | one `stats::lm()` per feature, full versus reduced model, no shrinkage |
| `msqrob` | `msqrob2`, aggregating features then fitting |
| `msqrob_agg` | `msqrob2::msqrobAggregate()`, fitting a gene's features together |
| `proda` | `proDA::proDA()`; natively handles missing values |
| `prolfqua` | `prolfqua::strategy_lm()` -> `build_model()` -> `get_anova()` |
| `prolfqua_lmer` | `prolfqua::strategy_lmer()`, one mixed model per gene over its features |
| `trend` | `limma::lmFit()` -> `eBayes(trend=TRUE)` -> `topTable()` |
| `voom` | `limma::voom()` -> `lmFit()` -> `eBayes(trend=FALSE)` -> `topTable()` |

`lm` is the unmoderated reference point the others are worth comparing against. `deqms`,
`msqrob`, `msqrob_agg` and `prolfqua_lmer` take feature-level input and report one row per
gene whatever level the input is at, so `combine_features` should be left out of
`config$run_order` for them. `config$test_moderate` and `config$test_trend` control shrinkage
of the per-feature error variance where the engine allows it; `config$test_random_obs` and
`config$test_ridge` are for the mixed-model methods.

Engines other than the `limma` ones need their package installed, so a portable script
guards on it:

```{r optional_engine}
if(requireNamespace("proDA", quietly=TRUE) &&
   requireNamespace("SummarizedExperiment", quietly=TRUE)) {
  res_proda <- h0testr::test_h0(out$state, out$config, method="proda")
  print(head(res_proda$standard, 3))
} else {
  cat("proDA not installed; skipping\n")
}
```

## Tuning by permutation

Which combination of normalization, imputation and testing suits a dataset is an empirical
question. `tune()` sweeps a grid of settings and counts hits for each combination; running it
again with a variable in `config$test_term` permuted gives the null those counts should be
read against.

In practice: one run with no permutation, then around 20 with permutation. The naive loop
below suits small to medium datasets; for larger ones, launch each instance separately in its
own directory. These chunks are not evaluated here, since a grid this size takes far longer
than a vignette should.

```{r tune, eval=FALSE}
grid <- list(
  normalization_methods=c("RLE", "q75", "cpm", "log2"),
  impute_methods=c("sample_lod", "unif_sample_lod", "none"),
  impute_quantiles=c(0, 0.05, 0.1),
  test_methods=c("trend", "msqrob", "proda", "prolfqua")
)

## one run on unpermuted data:
config$permute_var <- ""
set.seed(100)
out0 <- do.call(h0testr::tune, c(list(config=config), grid))
write.table(out0, "0.grp.tune.tsv", quote=FALSE, sep="\t", row.names=FALSE)

## then N runs with the tested variable permuted:
config$permute_var <- "grp"
for(idx in 1:20) {
  set.seed(100 + idx)
  outi <- do.call(h0testr::tune, c(list(config=config), grid))
  write.table(outi, paste0(idx, ".grp.tune.tsv"), quote=FALSE, sep="\t", row.names=FALSE)
}
```

Only a variable appearing in `config$test_term` is worth permuting. With `test_term="grp"`
that is `"grp"`; with `test_term="age:gender"` either `"age"` or `"gender"` would do.

Each run returns one row per grid point, with the settings that produced it (`norm`,
`nquant`, `impute`, `iquant`, `scale`, `span`, `npcs`, `k`, `test`, `perm`), the number of
tests with adjusted p-value below `0.05` (`nhits`), the number of tests performed (`ntests`),
and a completion timestamp (`time`).

`tune_check()` then collects the files and summarizes each grid point across the
permutations:

```{r tune_check, eval=FALSE}
tbl <- h0testr::tune_check(dir_in="/path/to/tuning/results", prefix="",
  suffix=".grp.tune.tsv", config=config)
```

| column | meaning |
|---|---|
| `nhits` | hits in the unpermuted run |
| `ntests` | tests performed in the unpermuted run |
| `fdr` | `max1 / nhits`, capped at `1`; `NA` when the combination never ran |
| `max1` | most hits in any one permutation |
| `mid1`, `avg1`, `sd1` | median, mean and SD of hits across permutations |

The remaining columns are the settings, as above. A grid point worth using has many `nhits`
and a low `fdr`; one with a large `max1` is a setting whose apparent hits are noise.

## Session info

```{r session_info}
sessionInfo()
```
